---
title: "Chicago crime/temperature example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chicago crime/temperature example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The data

To demonstrate the usefulness of EDM in estimating the impact of causal variables, we use a real-world dataset that reflects
daily temperature and crime levels in Chicago, which we make available in the [chicago.csv](chicago.csv) file.

First, we load the time series into R:

```{r load data}
chicagoURL <- url("https://github.com/EDM-Developers/fastEDM/raw/master/vignettes/chicago.csv")
chicago <- read.csv(chicagoURL)
head(chicago)
```

Plotting the two time series shows:

```{r plot time series (line plot)}
library(ggplot2)

df <- data.frame(list(
  t = rep(chicago$Time, 2),
  variable = c(rep("Crime", nrow(chicago)), rep("Temperature", nrow(chicago))),
  value = c(chicago$Crime, chicago$Temperature)))

ggplot(df, aes(x = t)) + 
  geom_line(aes(y = value, color = variable)) +
  labs(colour = "Time series")
```

When plotted as a scatter plot, there appears to be some linear correlation between the variables:

```{r plot time series (scatter plot)}
ggplot(chicago, aes(x = Temperature, y=Crime)) + geom_point(color="#F8766D")
```

Calculating the linear correlation between `Crime` and `Temperature`:

```{r linear correlation}
cor(chicago["Temperature"], chicago["Crime"])
```

shows a mild correlation, however the causal structure (if any) and its direction is not shown.

## Find the optimal embedding dimension

Now we use `edm` to find the optimal embedding dimension of the `Temperature` time series.
We check the values of $E = 2, \dots 20$.
The `crossfold=5` option means that, for each $E$ value we run 5 sets of predictions, and for each set we take four fifths of the data for training and predict the remaining one fifth.

```{r reduce dataset size}
# Just to speed up the example
chicago <- head(chicago, 500) # TODO: Remove this later.
```

```{r find optimal embedding dimension}
library(fastEDM)
edm(chicago["Time"], chicago["Temperature"], E=2:10, numThreads=4)
```

From the `rho` column we can see that the prediction accuracy is maximised when $E = 7$, so we take this as our estimate of the embedding dimension.

## Convergent cross-mapping

The `edm` command can also run the cross-mapping task, which allows us to ascertain the causal links between the crime and temperature time series.

```{r convergent cross mapping}
# libs <- c(seq(10, 200, 5), seq(210, 1000, 10), seq(1020, 2000, 20),
#           seq(2050, 4350, 50), 4365)

libs <- seq(10, 4300, 100)

tempPredictsCrime <- edm(chicago["Time"], chicago["Temperature"], chicago["Crime"],
                         algorithm="smap", E=7, library=libs, numReps=4, numThreads=4)

crimePredictsTemp <- edm(chicago["Time"], chicago["Crime"], chicago["Temperature"],
                         algorithm="smap", E=7, library=libs, numReps=4, numThreads=4)
```



!!! note
        This selects a lot of library points, and replicates the analysis some times, so this command may take a minute or two to finish.
        Choosing a machine with more CPU cores or faster cores will help significantly.   

Plotting the results gives:


```{r plot ccm rho-L graph}
library(tidyr)

averaged1 <- stats::aggregate(rho ~ library, tempPredictsCrime$summary, mean)
averaged2 <- stats::aggregate(rho ~ library, crimePredictsTemp$summary, mean)

averaged <- data.frame(Library = averaged1$library,
                       temp.to.crime = averaged1$rho,
                       crime.to.temp = averaged2$rho)

combined <- gather(averaged, key = "Direction", value = "Correlation", -Library)
combined[combined$Direction == "temp.to.crime", "Direction"] <- "Crime | M(Temperature)"
combined[combined$Direction == "crime.to.temp", "Direction"] <- "Temperature | M(Crime)"

ggplot(combined, aes(x = Library, y = Correlation)) + 
  geom_line(aes(color = Direction)) + scale_color_manual(values = c("darkred", "steelblue")) +
  geom_point(data = tempPredictsCrime$summary, aes(x = library, y = rho), alpha = 0.05, color = "darkred") +
  geom_point(data = crimePredictsTemp$summary, aes(x = library, y = rho), alpha = 0.05, color = "steelblue") +
  ylim(c(0.275, 0.625))

```

<!-- Convergent cross-mapping plot of accuracy against library size](chicago-rho-L.svg) -->

In this plot, we can see that one direction shows a significant increase in accuracy as $L$ increases, whereas the other direction is pretty flat.
The direction which increases the most is the $\texttt{Temperature} \mid M(\texttt{Crime})$ direction.
This notation means we used  $\texttt{Crime}$ to predict $\texttt{Temperature}$, and due to the backward nature of EDM means it refers to the causal link $\texttt{Temperature} \to M(\texttt{Crime})$.
Therefore, we'd conclude that there is a causal link from temperature to crime, though no link in the reverse direction (which would be implausible).  

## Inspecting the S-map coefficients 

If we run `edm` with the `saveSMAPCoeffs=TRUE` option, we can store the fitted S-map coefficients into variable which start with the prefix `beta`.

```{r save s-map coefficients}
# Library size should be 4365
tempPredictsCrime <- edm(chicago["Time"], chicago["Temperature"], chicago["Crime"],
                         algorithm="smap", E=7, k=Inf, saveSMAPCoeffs=TRUE, 
                         numThreads=4)

crimePredictsTemp <- edm(chicago["Time"], chicago["Crime"], chicago["Temperature"],
                         algorithm="smap", E=7, k=Inf, saveSMAPCoeffs=TRUE,
                         numThreads=4)

# . edm xmap temp crime, e(7) alg(smap) k(-1) savesmap(beta)
# Percent complete: 0...10...20...30...40...50...60...70...80...90...
# Percent complete: 0...10...20...30...40...50...60...70...80...90...
# 
# Empirical Dynamic Modelling
# Convergent Cross-mapping result for variables temp and crime
# --------------------------------------------------------------------------
#                  Mapping    Library size             rho             MAE
# --------------------------------------------------------------------------
#    crime ~ crime|M(temp)            4365          .46784          136.93 
#     temp ~ temp|M(crime)            4365          .54886          14.661 
# --------------------------------------------------------------------------
# Note: Number of neighbours (k) is set to 4364
# Note: The embedding dimension E is 7

```

Plotting them allows us to see the contemporaneous effect of temperature on crime.

```{r}
# . twoway (kdensity beta1_b1_rep1), xtitle("Contemporaneous effect of temperatur
# > e on crime") ytitle("Density")

```
<!-- # ![Contemporaneous effect of temperature on crime](chicago-crime1.svg) -->

```{r}
# . twoway (scatter beta1_b1_rep1 temp, xtitle("Temperature (Fahrenheit)") ytitle
# > ("Contemporaneous effect of temperature on crime") msize(small))(lpoly beta1_
# > b1_rep1 temp), legend(on order( 1 "Local coefficient" 2 "Local polynomial smo
# > othing"))

```
<!-- ![Contemporaneous effect of temperature on crime](chicago-crime2.svg) -->

